Practical Type Flattening

### 2020-01-08 paper/paper.tex
blur some types

\usepackage{hyperref}
\hypersetup
{
  bookmarks       = true,
  unicode         = true,
  pdftitle        = "Practical Type Blurring",
  pdfauthor       = "Ta Thanh Dinh", %auteur du document
%   pdfsubject      = "A Conceptual Approach for Message Format Extraction",
  pdftoolbar      = true, %barre d'outils non visible
  pdfmenubar      = true, %barre de menu visible
  pdfhighlight    = /O, %effet d'un clic sur un lien hypertexte
  colorlinks      = true, %couleurs sur les liens hypertextes
%   pdfpagemode = None, %aucun mode de page
%   pdfpagelayout = SinglePage, %ouverture en simple page
  pdffitwindow    = true, %pages ouvertes entierement dans toute la fenetre
  linkcolor       = red, %couleur des liens hypertextes internes
  citecolor       = green, %couleur des liens pour les citations
  urlcolor        = cyan, %couleur des liens pour les url
%   pagebackref=true
%   bookmarksdepth  = 3
}

@@

\begin{document}
%
\title{Practical Type Blurring}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ta Thanh Dinh}
%
\authorrunning{Ta Thanh Dinh}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
\email{tathanhdinh@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
High-level type reconstruction is an important step in machine code decompilation.
Researchers show possibilities of mapping untyped machine-dependent
primitives into types of a C-like type system.

We show that the techniques used for type reconstruction can be bypassed,
and we claim that the bypass is always possible given \emph{semantics gap} between
type systems of high-level and machine language.
% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
Binary code type inference is to restore high-level type information of variables,
functions from machine codes. It is useful for reverse code engineering (e.g.
decompilation) or static binary rewriting/instrumentation, just to name a few.
Type reconstruction is available in several binary analysis
tools~\cite{noauthor_hex-rays_nodate, noauthor_jeb_nodate, noauthor_ghidra_nodate}
as a result of machine code decompilation. Academic research includes~\cite{mycroft_type-based_1999}
which is one of the pilot papers on this domain, an improvement supporting
a limited form of subtyping is presented in~\cite{lee_tie_2011}, a survey of
research up until 2016 is given in~\cite{caballero_type_2016}. One of recent
directions is to use machine learning~\cite{maier_typeminer_2019}, which is out of scope of
the paper. All approaches tend to output a C-like type system.

Basically, the type inference starts by detecting high-level variables from
machine-dependent primitives: local variables are stored in function stack frame,
parameters are passed from registers, see~\cite{balakrishnan_wysinwyx_2007} for example.
Type constraints are generated then through the use of these variables. Finally,
the constraints are solved, each variable is annotated with a type given by the solver.

Though most papers agree that type information is removed in compilation~\cite{lee_tie_2011,caballero_type_2016,lin_automatic_2010},
there is no explicit explication of \emph{why and how} some type information is not needed
(so removed) in machine codes. A popular point of view that types should attach with
storage specific primitives (e.g. \texttt{int} is some \texttt{32-bit} signed integer
stored in a register or stack), though mostly correct, does not reveal the nature of
types. This is one of the reasons why some researchers unify data structure reverse
engineering with type reconstruction~\cite{caballero_type_2016, caballero_polyglot_2007}.
In some languages other than C, types can be zero-sized~\cite{noauthor_phantom_nodate}, another example is
\emph{regions}~\cite{grossman_region-based_2002} which have no storage imprint.

The binary type reconstruction is influenced by \emph{Algorithm W} used originally to
type check/inference for programs of Hindley-Milner's type system (abbr. HM)~\cite{milner_theory_1978,hindley_principal_1969,cardelli_basic_1987}.
While the supposed target is a C-like type system, some authors claim to replicate
the result of \emph{principal types}~\cite{damas_principal_1982,hindley_principal_1969}
by recovering \emph{the most precise yet conservative} type~\cite{lee_tie_2011}.
Such a result is sound in a limited context only, C's type system is not HM: there
are functions that are trivially typed in C by casting, but not typeable in HM's.
Actually, it is not rare to observe type inconsistency in decompilation results
of security tools.

\paragraph{Contribution}
We first discuss the type information loss in compilation and some cases
where assigning a high-level \emph{type} for a low-level primitive
is not possible. Though this part is not novel, we bring it to the context
of binary code analysis to show limits of type reconstruction. Next, we present
a C compiler which tries to hide types from reconstruction techniques described
in current researches.

\section{Type information loss}
Type system of a language can be considered as a set or rules about constraints
on programs of this language. On statically typed languages, the constraints
can be mechanically checked and proved without running the program, it helps
avoid bugs as Robin Milner's famous slogan~\textquote{well-type program cannot go wrong}~\cite{milner_theory_1978}.

There exists at least two sources information loss: \emph{type erasure} and
\emph{data indistinguishability}~\cite{noauthor_struct_nodate}, happening at
different phases of compilation.

\subsection{Type erasure}
Once the program is type checked, the type information is normally not needed for
evaluation (i.e. running program): the compiler can select a consistent machine-dependent
representation for each high-level type, this representation may only have little
relation with the original type but it does not violate the fact that the program
is checked (then safe).

>>>>

\usepackage{hyperref}
\hypersetup
{
  bookmarks       = true,
  unicode         = true,
  pdftitle        = "Practical Type Flattening",
  pdfauthor       = "Ta Thanh Dinh", %auteur du document
%   pdfsubject      = "A Conceptual Approach for Message Format Extraction",
  pdftoolbar      = true, %barre d'outils non visible
  pdfmenubar      = true, %barre de menu visible
  pdfhighlight    = /O, %effet d'un clic sur un lien hypertexte
  colorlinks      = true, %couleurs sur les liens hypertextes
%   pdfpagemode = None, %aucun mode de page
%   pdfpagelayout = SinglePage, %ouverture en simple page
  pdffitwindow    = true, %pages ouvertes entierement dans toute la fenetre
  linkcolor       = red, %couleur des liens hypertextes internes
  citecolor       = green, %couleur des liens pour les citations
  urlcolor        = cyan, %couleur des liens pour les url
%   pagebackref=true
%   bookmarksdepth  = 3
}

@@

\begin{document}
%
\title{Practical Type Flattening}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ta Thanh Dinh}
%
\authorrunning{Ta Thanh Dinh}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
\email{tathanhdinh@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
High-level type reconstruction is an important step in machine code decompilation.
Researchers show possibilities of mapping untyped machine-dependent
primitives into types of a C-like type system.

We show that the techniques used for type reconstruction can be bypassed,
and we claim that the bypass is always possible given \emph{semantics gap} between
type systems of high-level and machine language.
% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
Binary code type recovery is to restore high-level type information of variables,
functions from machine codes. It is useful for reverse code engineering (e.g.
decompilation) or static binary rewriting/instrumentation, just to name a few.
Type recovery is available in several binary analysis
tools~\cite{noauthor_hex-rays_nodate, noauthor_jeb_nodate, noauthor_ghidra_nodate}
as a result of machine code decompilation. Academic research includes the classic
paper of Mycroft~\cite{mycroft_type-based_1999} which is probably the first work
on the domain, though earlier ideas~\cite{shivers_data-flow_1990} have been used
to recover types of programs in Scheme - a dynamically typed language.
Lee et al.~\cite{lee_tie_2011} introduces
an extension for Mycroft's work which supports a limited form of subtyping.
Caballero and Lin give a survey~\cite{caballero_type_2016} for research up until 2015.
Recent directions are using machine learning~\cite{maier_typeminer_2019} or statistical
language model~\cite{katz_estimating_2016}, which are out of scope of the paper.
All approaches tend to output a C-like type system.

Basically, the binary code type recovery starts by detecting high-level objects from
machine-dependent primitives: local variables are stored in function stack frame,
parameters are passed from registers, see~\cite{balakrishnan_wysinwyx_2007} for related techniques.
Type constraints are generated gradually through the \empĥ{low-level use} of objects.
The system of constraints is solved then, each object is assigned a certain
(polymorphic) type.

This bird view may give off an impression that the
binary code type reconstruction is a well established research domain, incorrect results
are because of foreign reasons: incorrect binary code disassembling,
high-level variable detection, etc. But we claim that there are probably intrinsic problems.

\emph{First}, there is a common agreement that type information is removed through
compilation~\cite{lee_tie_2011,caballero_type_2016,lin_automatic_2010},
but there is no explicit explication of why and how some type information is not needed,
so removed, in machine codes. The result of reconstructing \emph{the most precise yet conservative}
type~\cite{lee_tie_2011}, which may suggest the well-known
\emph{principal types}~\cite{damas_principal_1982,hindley_principal_1969},
is sound in a limited context only. Actually not all type
information can be restored.

\emph{Second}, though some techniques of data structure reverse
engineering~\cite{caballero_type_2016, caballero_polyglot_2007} can be reused for
type recovery, types do not always attach with storage specific primitives (for example
\texttt{int} is some \texttt{32-bit} signed integer stored in a register or stack).
The storage-based point of view is almost correct in low-level languages like C, it
does not reveal the nature of types. In other languages, types can be
zero-sized~\cite{noauthor_phantom_nodate}, another examples are
\emph{regions}~\cite{grossman_region-based_2002} or \emph{units-of-measure}~\cite{kennedy_types_2010}
which leave no storage imprint.

\emph{Third}, current research uses some variants of
\emph{Algorithm W}~\cite{milner_theory_1978,cardelli_basic_1987} to resolve type
constraints in a C-like language. But the polymorphism in C is rather
\emph{ad-hoc}~\cite{strachey_fundamental_2000}: some functions are
trivially typed in C by casting, but not typeable in Hindley-Milner type system. This
problem is remarked already in~\cite{mycroft_type-based_1999}, but unfortunately omitted in
later extension~\cite{lee_tie_2011}. In real world, it is not rare to see type inconsistency in decompilation results
of security tools.

% used originally to
% type check/inference for programs of Hindley-Milner's type system (abbr. HM).
% While the supposed target is a C-like type system, some authors claim to replicate
% the result of \emph{principal types}~\cite{damas_principal_1982,hindley_principal_1969}
% by recovering \emph{the most precise yet conservative} type~\cite{lee_tie_2011}.
% Such a result is sound in a limited context only, C's type system is not HM: there
% are functions that are trivially typed in C by casting, but not typeable in HM's.
% Actually, it is not rare to observe type inconsistency in decompilation results
% of security tools.

\paragraph{Contribution}
We first discuss the type information loss in compilation and some cases
where assigning a high-level \emph{type} for a low-level primitive
is not possible. Though this part is not novel, we bring it to the context
of binary code analysis to show limits of type reconstruction. Next, we present
a proof-of-concept C compiler which tries to hide types from reconstruction techniques described
in current researches.

\section{Type information loss}
Informally speaking, a type system is some set of constraints and rules to infer constraints
on a program. The type system decides which are allowed or forbidden on the semantics of the language,
for example adding two pointers is not allowed in C because this operation is semantically nonsense.
Type helps avoid bugs as summarized in Milner's famous
slogan~\textquote{well-type program cannot go wrong}~\cite{milner_theory_1978}.

In statically typed languages, the compiler checks the type constraints then generates
machine codes (we consider here only languages which generates machine codes), but then
most of constraints are lost because they are not needed for running
the program. There are at least two sources of
information loss: \emph{type erasure} and \emph{data indistinguishability},
happening at different phases of compilation.

\subsection{Type erasure}
\emph{Type erasure} means the "intensional" type of an object is hidden/inaccessible or
completely removed in some context. It appears under several concepts: subtype substitution,
generic specialization (or monomorphization), or boxing, just name a few. In some cases,
e.g. subtype substitution, the (sub)type of the wrapped object does not completely removed,
it is just not accessible in contexts where only behaviors of the super type are allowed.
Such a case will not be discussed in the paper since we look at the low-level machine code.

Once the program is type checked,
% the type information is normally not needed for evaluation (i.e. running program),
 the compiler will select a consistent machine-dependent
representation for each high-level type, this representation may only have little
relation with the original type but it does not violate the fact that the program
is checked, then safe.
